---
title: "Obtain a non-redundant set of AA indices for given sequences"
---

Generate all AAindex values for given set of sequences,
then keep only uncorrelated AA indices.

Include also KYTJ820101 Hydropathy index (Kyte-Doolittle, 1982) since it 
appears frequently in literature + we want to make sure to have this type of property.


```{r, setup, message=F}
library("bio3d")
library("rhdf5")
library("tidyverse")
library("Biostrings")
library("progress")
```

```{r}
prot_seq <- Biostrings::readAAStringSet(
    "../DeepET_reps/data/cleaned_enzyme_topts_with_structures.fasta"
    )

prot_seq <- tibble(
    uniprot_ac = names(prot_seq),
    seq = paste(prot_seq)
    ) %>% 
    dplyr::mutate(uniprot_ac = base::sapply(
        uniprot_ac,
        function(id_str) {
           stringr::str_split(id_str, pattern = "\\;")[[1]][[1]]
        })
    )
```

## Get AAindex profiles

Get values from all AA indices (release 9.1 2006), then remove redundant indices
(based on a subsample of 1000 sequences, for the sake of memory and time)

```{r}
data(aa.index)
N_SEQ_SUBSAMPLE = 1000

#' For the given protein sequence, 
#' return correlation matrix of all AAindices calculated on the sequence.
#' NOTE: Assumes aa.index loaded into namespace, i.e. \code{data(aa.index)}
aaindex_corr_matrix_for_seq <- function(prot_seq) {
    index_values_for_seq <- base::sapply(
        names(aa.index),
        function(aa_idx) { 
            aa2index(str_split(prot_seq, pattern = "")[[1]], index = aa_idx, window = 1) }
    )
    return(cor(index_values_for_seq, use = "complete.obs"))
}

aaindex_corr_matrices <- 
    lapply(
        prot_seq %>% dplyr::sample_n(N_SEQ_SUBSAMPLE) %>% select(seq) %>% pull,
        aaindex_corr_matrix_for_seq
        )

print(length(aaindex_corr_matrices))
aaindex_avg_corr_matrix <- abs(base::Reduce("+", aaindex_corr_matrices) / N_SEQ_SUBSAMPLE)

# Clean up large list of matrices
rm(aaindex_corr_matrices)
gc()
```

```{r}
reshape2::melt(na.omit(aaindex_avg_corr_matrix)) %>% 
ggplot(aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() +
  scale_fill_gradient2(low = "white", high = "black", na.value = "pink") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

```{r, message=F}
library("caret")
correlated_indices = sort(caret::findCorrelation(na.omit(aaindex_avg_corr_matrix),
                                                 cutoff = 0.5))
uncorr_indices <- base::colnames(aaindex_avg_corr_matrix[-c(correlated_indices),
                                                         -c(correlated_indices)])
uncorr_indices <- c(uncorr_indices, "KYTJ820101")
```

```{r}
uncorr_indices
```

```{r}
reshape2::melt(na.omit(aaindex_avg_corr_matrix[c(uncorr_indices),
                                               c(uncorr_indices)])) %>% 
ggplot(aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() +
  scale_fill_gradient2(low = "white", high = "black", na.value = "pink") +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```


```{r}
prog_bar <- progress_bar$new(total = length(uncorr_indices))

for (aa_index in uncorr_indices) {
    prog_bar$tick()
    
    prot_seq[[aa_index]] = base::sapply(
        prot_seq[["seq"]],
        function(seq) {
            aa2index(str_split(seq, pattern = "")[[1]], 
                     index = aa_index, 
                     window = 1)
        }
    )
}

prot_profiles <- prot_seq %>% select(-seq)
```


```{r}
protein_ids <- prot_profiles %>% select(uniprot_ac) %>% pull
```


```{r}
store_file <- "data/protein_features/uncorr_aaindex_profiles.h5"
if (!h5createFile(store_file)){
    stop("Could not create H5 store")

} else {
    prog_bar <- progress_bar$new(total = length(protein_ids))
    for (prot_id in protein_ids) {
        prog_bar$tick()
        
        prot_data <- prot_profiles %>% dplyr::filter(uniprot_ac == prot_id)
        
        h5createGroup(store_file, prot_id)
        
        for (aa_index in uncorr_indices) {
            index_values <- prot_data %>% dplyr::select(all_of(aa_index)) %>% pull
            attr(index_values, "names") <- NULL
            h5write(index_values, store_file, base::paste(prot_id, aa_index, sep = "/"))
        }
    }
    
}
```


